---
title: "Intro to Data Visualization"
author: "Your Name Here"
date: "October __, 2023"
editor: visual
format: 
  html:
    code-overflow: wrap
---

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggbeeswarm)

```

## Import the Water Quality Data

Run the code below to import the water quality data

```{r}
#| label: import data

waterQ <- read_csv("waterQ.csv",
          col_types = cols(
            date = col_date(format = "%Y-%m-%d"), 
            year = col_factor(),
            season = col_factor(levels = c("Summer", "Fall")),
            location = col_factor(levels = c("Downstream","Upstream")),
            microhabitat = col_factor(
              levels = 
                c("DSR", "DFR", "DM", "DSH", "DSP",
                  "USR", "UFR", "UM", "USH", "USU"))
                            ))

```

Use the skills you learned in the "Getting Data into R" tutorial to look at the structure of the dataframe that was just imported.

I personally usually click on the name of the dataframe in the Global Environment as a shortcut to using the View() function, and click on the blue arrow next to the dataframe in the Global Environment to see all the variable names and types

Also click on the Files tab and click on the file called "waterQ_metadata.txt" to see a full description of the data including how to cite the data, what each variable represents, and how measurements were made using what tools.

## How does Paradise Pond affect water quality in the Mill River?

We can use this dataset to look at differences in the water quality upstream versus downstream of Paradise Pond to test for possible effects of the pond on the Mill River.

Keep in mind that Paradise Pond can have effects on the downstream environment beyond sediment redistribution. Sediment redistribution that occurred during the collection of this water quality dataset was in the winter (November-January) of most years starting in 2019. In today's lab, we will be concentrating more on the general effect of the pond rather than from sediment redistribution.

Paradise Pond is an impoundment in the river that slows the flow and makes it more like a pond and less like a river. The water falling over the dam creates a lot of turbulence. How might the slow flow in the pond and/or the turbulence from the water falling over the dam affect the water quality downstream vs upstream?

**Question: Look at the water quality variables and choose ONE that you want to explore today. Type below which variable you chose:**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

**Question: What do you predict to see in the data comparing the upstream and downstream locations for this variable? (in other words, How do you predict Paradise Pond will affect this variable (if at all)?). Write your prediction and explain why you made that prediction below:**

**For help understanding the variable better and how the pond might affect it, see the [Background Information on Water Quality Measurements](https://docs.google.com/document/d/1j0aGgQS0CMSPLzv-2JTGoWulUJEeRxhiMKhggGTxysU/edit?usp=sharing)**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

## Missing Data

Does your variable of interest have missing data? Replace the blank in the code below with the column name of your variable of interest to summarize that variable.

Make sure you spell and capitalize the variable name EXACTLY how it is in the waterQ data frame.

```{r}
#| label: Summary of Data

summary(waterQ$____) #replace the blank with the column name of your variable

```

**Question: How many NA's (=missing data) does your variable of interest have?**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

Let's remove the NA's for your variable. Run the code below making sure to replace the blank below with the name of your variable of interest (again, use the exact match to the column name of the variable of interest)

```{r}
#| label: remove NAs

waterQ2 <- waterQ |> 
  filter(!is.na(_____))

```

While this step is not strictly necessary, let's remove some of the columns we don't need to make the data easier to look at. Run the code below making sure to replace the blank below with the name of your variable of interest.

```{r}
#| label: remove unneeded columns

waterQ2 <- waterQ2 |> 
  select(date, year, season, location, sampleID, sampling, microhabitat, quadrat, _______)

```

Look at the new dataset (waterQ2) in the Global Environment. The new number of observations should be the original number minus the number of NA's. The number of variables (=columns) should be 9.

## Descriptive Statistics

Next let's calculate the descriptive statistics for your variable of interest for the Mill River water quality data. Run the code below making sure to replace the blanks below with the name of your variable of interest. (note - do not place anything in the empty parentheses in the n(), only replace the underlined blanks for the mean, median, SD and IQR)

```{r}
#| label: descriptive stats

waterQ2 |>                        #name of the data frame is here
  group_by(location) |>           #name of grouping variable (=factor) is here
  summarize(mean = mean(___),     #water quality variable name here
            median = median(___), #water quality variable name here
            SD = sd(___),         #water quality variable name here
            IQR = IQR(___),       #water quality variable name here
            n())                  #don't put anything here, leave as is

```

How do we report these values? We have to round appropriately and include the correct units.

**Question: What units is your variable of interest measured in? (Look at the "waterQ_metadata.txt" file in the dataviz folder and then scroll down to your variable of interest to see what units it was measured in)**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

How should we round these values?

Note, you can look at the "waterQ_metadata.txt" file in the data folder and then scroll down to your variable of interest to see that level of accuracy the variable was measured to. You can usually go one digit to the right beyond the level of accuracy (because that next digit is the estimated digit) at most. Note that it is okay to be a little more conservative and round more, you just don't want to give the impression that you have more accuracy than you do by giving too many digits.

Personally, I think it is easiest to use the standard deviation (SD) to tell us on average how much variability there is in the data.

Start by rounding the standard deviation (SD) to 2 SIGNIFICANT DIGITS. (Look above for the descriptive results you already calculated, the SD for each location is given there)

**Question: Give the rounded values for SD (use 2 significant digits) for each group below:**

::: {.callout-note appearance="minimal" icon="false"}
Downstream SD:

Upstream SD:
:::

To round the mean, you round it to the same DECIMAL place as the SD. Rounding the SD tells us how variable the data are and how many decimals we can report to.

**Question: Give the rounded values for mean for each group below:**

::: {.callout-note appearance="minimal" icon="false"}
Downstream mean:

Upstream mean:
:::

Round the median and interquartile range (IQR) the same way you did for the mean.

**Question: Give the rounded values for median and IQR for each group below:**

::: {.callout-note appearance="minimal" icon="false"}
Downstream median:

Downstream IQR:

Upstream median:

Upstream IQR:
:::

**Question: Are the values for the mean and median within each group similar or different? (compare the downstream mean to the downstream median, then compare the upstream mean to the upstream median. You have to use some judgment about what is "similar" vs "different" at this point)**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

**Question: Can you tell at this point if the data for your variable of interest is normally distributed or not?**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

## Data Visualization

Next we are going to visualize the data. Before you do this for the Mill River water quality data, let's first get some practice by doing the "Graphing with ggplot" lesson from "Intro to R for Biology".

IT IS A GOOD IDEA TO SAVE THIS .qmd FILE BEFORE YOU START THE TUTORIAL IN CASE YOU HAVE ANY PROBLEMS!!

Complete the tutorial called [Intro to R for Biology: Graphing with ggplot](https://shiny.smith.edu:3838/mcpratt/r4bio/ggplot/)

(remember to use the [Smith College VPN](https://www.smith.edu/its/tara/networking/vpn.html) if you are off campus to access the tutorial)

### Histogram

Let's first use a histogram to look at the distribution of your water quality variable of interest for each group. (just note that we are graphing these histograms sideways compared to how they are typically graphed to keep things consistent with the having the water quality variable as the y-variable)

The code to make histograms for each of the water quality variables is included in the code chunk below. Notice that the only thing that you need to change is the **y-variable** (must match column name for your water quality variable from dataframe), the **binwidth**, and the **y-label** (the nicer label that will show up on the figure). Place the name of the column of your water quality variable in the first blank, play with the binwidth to give you enough but not too many bins (something similar to the SD you got above might be a good starting place), and replace the ylab line using the following code below as needed:

```         
ylab("Flow Rate (m/s)") +

ylab("Dissolved Oxygen (mg/L)") +

ylab(~paste("PAR"," ", "(" ,mu, "mol m"^-2,"s"^-1, ")")) +

ylab("pH") +

ylab("Water Temperature ("~degree~"C)") +

ylab("Turbidity (NTU)") +

ylab(~paste("Conductivity"," ", "(" ,mu, "S cm"^-1, ")")) +

ylab("Nitrate Concentration (mg/L)") +

ylab(expression(Alkalinity~(ppm~of~CaCO[3]))) +

ylab(expression(Phosphate~Concentration~(ppm~of~PO[4]^{"3-"}))) +
```

```{r}
#| label: Histogram

#Histogram  
ggplot(data = waterQ2, 
       aes(y = __, fill=location)) + #y = water quality variable
  geom_histogram(binwidth = __,   #adjust the binwidth as needed
                 color = "white", show.legend = FALSE) + 
  facet_grid(location ~ .) +
  ylab("_____") + #replace this line with appropriate one above
  xlab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20) #this number can be decreased if labels get cut off

```

**Question: Does the distribution for each location look like a normal distribution?**

::: {.callout-note appearance="minimal" icon="false"}
\[Delete this text and place your answer here\]
:::

### Choosing the Best Graph

In the final activity for this assignment, you will make a graph that shows the water quality variable you chose for both locations. Think about what type of variables each is (what type of variable is the water quality variable? what type of variable is location?), and then which one should be the dependent variable (on the y-axis) and which should be the independent variable (on the x-axis). We want to know what effect the independent variable has on the dependent variable.

Given what you have learned above (what is the sample size in each group? is your water quality variable normally distributed?), think about what would be the best way to represent the water quality variable Upstream vs Downstream of Paradise Pond in the Mill River. Remember that it is a best practice to include something that shows the distribution of your numerical variables and include a measure of central tendency (mean and/or median as appropriate). You also want to give a measure of dispersion (IQR, SD, range, etc) or show all the points. Take a look at this link to remind you of some of the best practices in [choosing a graph type](https://docs.google.com/presentation/d/1qSSHUw_SIXtrRFIxIqa4IfHOwUJetSD9qi6GUC7GsoE/edit?usp=sharing).

Finally, make sure you give an indication of the sample size in each group by either including that on the graph itself (this is more advanced code, so I recommend that you only chose this if you already have some experience graphing in ggplot2 and want to push yourself to learn something more advanced), or by writing what the sample size is in the figure legend below the graph (you will write the figure legend in a Google Doc). Recall that you had R calculate descriptive statistics including the sample size for each group above.

I have created a bunch of scripts with a template you can fill in the blanks for to make your own version of what you think is the best graphing option. Open the "script_templates" folder and look through the script files for the code you want to try. Copy the code of the graph type you think would be best and paste the code in the code chunk below (feel free to try several before you decide which one is best).

Tweak the script template code so that it works with your water quality variable of interest. Feel free to play around to see what happens when you remove or change things in the code.

```{r}
#| label: best graph




```

## Exporting your best graph

How do you save a copy of your graph to use elsewhere?

The simplest thing to do is probably just right-click (Windows) or control-click (Mac) on the graph and select "Copy Image" and then just paste the image where you want it. However, if you want a higher quality image where you can specify the size, image quality, and file type you can use the `ggsave()` function.

When saving with the `ggsave()` function, you need to name the resulting file with surrounding " ", and indicate the size via height, width, and units. Don't forget to save the graph with a dpi call between 300-500 to make it nice and crisp! Look at the `ggsave()` help file for more information and options.

```{r}
#| label: Save your graph
#| eval: false

# save the graph!
ggsave(filename="___.png",  #recommended to use.png or .jpg file types
       height = 7, width = 8, units = "in", 
       dpi = 300)

```

## Render this File to HTML then Print to PDF

You are all done with this activity, but you want to make sure to save all your work and put a copy of it in your OneNote lab notebook. Ultimately, you need a PDF file to put into your lab notebook. However, I have found that the formatting works best if you render to HTML first, then open that file with your browser and print to PDF from your browser.

Make sure the PDF file is saved somewhere on your computer that you can get to (you will first need to export this file to your computer if you are working on the Smith Server). Then insert the PDF of this file into your lab notebook as instructed.
