---
title: "effectsize_NAME"
subtitle: "BIO 131 Intro to Effect Size"
author: "Your Name Here"
date: "Mar , 2025"
format: typst
---

Before you start working, give this file a new name so it doesn't get written over if you accidentally download the GitHub repository a second time. Go to File in the menu and choose "Save as" and then give it the name "effectsize_Name.qmd" (replace "Name" with your full name)

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse) ## for readr, dplyr, ggplot2
library(rstatix) ## for effect sizes

```

## Import the Water Quality Data

Run the code below to import the water quality data

```{r}
#| label: import data

waterQ <- read_csv("data/waterQ.csv",
          col_types = cols(
            date = col_date(format = "%Y-%m-%d"), 
            year = col_factor(),
            season = col_factor(levels = c("Summer", "Fall")),
            location = col_factor(levels = c("Downstream","Upstream")),
            microhabitat = col_factor(
              levels = 
                c("DSR", "DFR", "DM", "DSH", "DSP",
                  "USR", "UFR", "UM", "USH", "USU"))
                            ))

metadata <- read_csv("data/waterQ_variable_metadata.csv")

#data cleaning to remove missing values etc
#replace the blanks with the column name of your variable
waterQ2 <- waterQ |> 
  filter(!is.na(_____)) |> 
  select(date, year, location, ___)

```

## Inference - Effect Size

How do we know if there is a difference for this water quality variable between the locations upstream and downstream of Paradise Pond? We use inferential statistics to help us determine whether a sample can help tell us about the full population and make comparisons. In this case, we want to compare between two groups (upstream vs downstream).

You will learn much more about inferential statistics in a full statistics course. In this class, we are just going to discuss effect sizes. An effect size tells us how big the difference is between groups while taking into account the variability as well.

To choose the right effect size, you need to take the distribution and variance into account. Look back at your QQ plot from the last lab and remind yourself if the data is normally distributed or not. Look at your descriptive statistics and graph to see if the dispersion (SD, IQR, range, etc) looks similar or not between the two locations.

-   **Normal Distribution, Equal Variance:** If the data is normal, and the two groups have similar dispersion (i.e. similar variance), then you should use "regular" **Cohen's d effect size (*d*)**

-   **Normal Distribution, Unequal Variance:** If the data is normal, but the two groups have unequal dispersion (i.e. unequal variance), then you should use the **Cohen's d for Welch test effect size (*d*)**

-   **Non-normal Distribution:** If the data is NOT normal, then you should use the **Wilcoxon effect size (*r*)** (also note that some statisticians recommend to use the Cohen's d for Welch test effect size if the variance is very different even if the distribution is not normal)

Choose the correct effect size test for your data using the criteria listed above and then copy and paste the corresponding code from the "inference" subfolder in the "script_templates" folder and paste it below then fill in the blanks as directed.

Note the following when filling in the blanks

-   data frame name = waterQ2

-   dependent variable = the water quality variable column name

-   factor = the independent categorical variable column name (in this case, location)

```{r}
#| label: effect size   





```

## Add Effect Size Results to Water Quality Results

From the effect size test output, write the following down in your lab notebook so you will have it all to include in your Water Quality Results:

-   what type of effect size you calculated (Cohen's d, Cohen's d for Welch test, or Wilcoxon effect size)

-   effect size (labeled **effsize** in the results table, round this to 1-2 decimal places)

-   95% confidence interval (labeled **conf.low** & **conf.high**, round this to 1-2 decimal places)

-   magnitude of the effect (labeled **magnitude**)

-   sample size for each location (labeled **n1** & **n2** - make sure you also note which location is group1 and which is group2 so you know which sample size goes with which location)

The effect size is standardized by the amount of variation in the data (the exact equation depends on which effect size you use). The interpretation of the standardized effect size is included in the magnitude column - negligible, small, moderate, or large. This will help you determine if the difference between means or medians is meaningful or not. The 95% confidence intervals tells you how precisely the effect size was estimated. If the 95% confidence interval does not include zero, it suggests that the effect is statistically significant

::: {.callout-note appearance="minimal"}
## Optional Additional Resources on Effect Sizes

-   How to use the `rstatix` package to calculate effect sizes

    -   **Cohen’s d for Student t-test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Cohen’s d for Welch test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Effect size** under **Wilcoxon rank sum test** on the [Wilcoxon Test in R](https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/) webpage

-   To learn more about inferential tests to compare two groups

    -   compare means with normal distribution: [t-Test](https://datatab.net/tutorial/t-test)

    -   rank-sum test to compare two independent samples that are not normally distributed: [Mann-Whitney U-Test](https://datatab.net/tutorial/mann-whitney-u-test)

    -   signed-rank test to compare paired groups that are not normally distributed: [Wilcoxon signed-rank test](https://datatab.net/tutorial/wilcoxon-test)

-   [Alternatives to P value: confidence interval and effect size](https://pmc.ncbi.nlm.nih.gov/articles/PMC5133225/)
:::

## Render this File to PDF

You are now all done with this activity, but you want to make sure to save all your work and put a copy of it in your Individual Google Drive folder for this class (along side your lab notebook). Ultimately, you need a PDF file to put into your lab notebook.

-   Before you Render

    -   Make sure that each code chunk has an unique label name or no name

    -   Save this file (make sure this file is named "effectsize_YourName.qmd" (replace "YourName" with your full name)

-   Render this Quarto file to PDF (just select the Render menu item at the top of this window)

-   Make sure the PDF file is saved somewhere on your computer that you can get to (you may need to export this file to your computer if you are working on the Smith server)

-   Place a copy of the PDF file in your INDIVIDUAL Google Drive folder for this class
