---
title: "effectsize_YourNAME"
subtitle: "BIO 131 Intro to Effect Size"
author: "Your Name Here"
date: "March , 2026"
format: typst
---

Before you start working, give this file a new name so it doesn't get written over if you accidentally download the GitHub repository a second time. Go to File in the menu and choose "Save as" and then give it the name "effectsize_YourName.qmd" (replace "YourName" with your full name without any spaces)

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse) ## for readr, dplyr, ggplot2
library(ggbeeswarm) ## for jittering points in a dotplot
library(ggpubr) ## for making QQ-plots and changing the color palette
library(rstatix) ## for effect sizes
library(patchwork)  ## for joining graphs

```

## Import the Water Quality Data

Run the code below to import the water quality data

```{r}
#| label: import data

waterQ <- read_csv("data/waterQ.csv",
          col_types = cols(
            date = col_date(format = "%Y-%m-%d"), 
            year = col_factor(),
            season = col_factor
            (levels = c("Winter","Spring", "Summer", "Fall")),
            location = col_factor
            (levels = c("Downstream","Upstream")),
                                        )) |> 
  filter(season != "Winter", season != "Spring")

metadata <- read_csv("data/waterQ_variable_metadata.csv")

```

Replace both of the blanks below with the column name of the water quality variable you worked with last week

```{r}
#| label: data cleaning

#data cleaning to remove missing values etc
#replace the blanks with the column name of your variable
waterQ2 <- waterQ |> 
  filter(!is.na(___)) |> 
  select(date, year, location, ___)

```

## Re-graph Water Quality Data

We are going to add some statistical results to your graph. Thus, first we need the code that created your graph from last time added into THIS Quarto document.

Go back to your previous Quarto document ("dataviz_YourName.qmd") and copy the code from the `best graph` code chunk and paste that below

```{r}
#| label: best graph




```

Go back to your previous Quarto document ("dataviz_YourName.qmd") and copy the code from the `best graph with accessible colors` code chunk and paste that below

```{r}
#| label: best graph with accessible colors



```

## Inference - Effect Size

How do we know if there is a difference for this water quality variable between the locations upstream and downstream of Paradise Pond? We use inferential statistics to help us determine whether a sample can help tell us about the full population and make comparisons. In this case, we want to compare between two groups (upstream vs downstream).

You will learn much more about inferential statistics in a full statistics course. In this class, we are just going to discuss effect sizes. An effect size tells us how big the difference is between groups while also considering the amount of variability.

To choose the right effect size, you need to take the distribution and variance into account. Remind yourself about whether the data is normally distributed or not for your water quality variable (HINT: look at your graph and see if it looks normal. Alternatively, go to the PDF of last lab's activities that is in your Individual Google Drive Folder for this class and look at your QQ plot). We also need to see if the variance is equal. Since the standard deviation is the square root of the variance, you can look at your descriptive statistics (also found in the PDF from the last lab) to see if the SD looks similar or not between the two locations.

-   **Normal Distribution, Equal Variance:** If the data is normal, and the two groups have similar dispersion (i.e. similar variance), then you should use the "regular" **Cohen's d effect size (*d*)**

-   **Unequal Variance:** If the data has unequal dispersion (i.e. unequal variance), then you should use the **Cohen's d for Welch test effect size (*d*)**

-   **Non-normal Distribution, Equal Variance:** If the data is NOT normal but the two groups have similar dispersion (i.e. similar variance), then you should use the **Wilcoxon effect size (*r*)** (note that some statisticians recommend to use the Cohen's d for Welch test effect size if the variance is very different even if the distribution is not normal)

Choose the correct effect size test for your data using the criteria listed above and then copy and paste the corresponding code from the "inference" subfolder in the "script_templates" folder and paste it below then fill in the blanks as directed.

Note the following when filling in the blanks

-   data frame name = waterQ2

-   dependent variable = the water quality variable column name

-   factor = the independent categorical variable column name (in this case, location)

```{r}
#| label: effect size   







```

## Record Effect Size Results

From the effect size test output, write the following down in your lab notebook so you will have it all to include in your Water Quality Results:

-   what type of effect size you calculated (Cohen's d, Cohen's d for Welch test, or Wilcoxon effect size)

-   effect size (labeled **effsize** in the results table, round this to 2 significant figures)

-   95% confidence interval (labeled **conf.low** & **conf.high**, round this to 2 significant figures)

-   magnitude of the effect (labeled **magnitude**)

-   sample size for each location (labeled **n1** & **n2** - make sure you also note which location is group1 and which is group2 so you know which sample size goes with which location)

The effect size is the difference between typical values standardized by the amount of variation in the data (the exact equation depends on which effect size you use). The 95% confidence interval was calculated by using the adjusted bootstrap percentile (BCa) method with 1000 bootstrapped samples.

The interpretation of the standardized effect size is included in the magnitude column - negligible, small, moderate, or large. This will help you determine if the difference between means or medians is meaningful or not. The 95% confidence intervals tells you how precisely the effect size was estimated. If the 95% confidence interval does not include zero, it suggests that the effect is statistically significant (if conf.low and conf.high are both positive or both negative than that means the 95% CI does not include zero. But if one is positive and the other is negative then that means the 95% CI does include zero).

## Graph Effect Size Results

Graphing the results of the effect size test makes it easier to understand and interpret. Thus, we will next graph your standardized effect size and the bootstrapped 95% confidence interval that you just calculated above. You don't need to change anything in the code below, just run it.

```{r}
#| label: effect size graph

#graph the results to make it easier to visualize
effect.plot <- effectSize.results  |>    
  ggplot(aes(
    x = .y., 
    y = effsize)) +
  
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  size = 1, # Adjust point size
                  linewidth = 0.8, # Adjust line thickness
                  color = "forestgreen") + # Set color
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") + # Add a reference line at 0
  
  #nicer axis labels
  xlab("") +
  ylab("Standardized Effect Size") +
  
  #to set the limits of the y-axis
  #you can replace these NA values as needed
  coord_cartesian(ylim = c(NA, NA)) +
  
  #additional formatting
  theme_classic(base_size = 14)  +  #sets the font size
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "none")   #controls legend/key

effect.plot


```

In your written resutls, make sure to talk about the magnitude of the effect (negligible, small, moderate, large? See `effectSize.results` output from the previous code chunk) and whether the effect was significantly different from zero or not (using the 95% CI). Does the 95% CI cross the dashed zero line? If the 95% CI includes zero, that means you cannot say that the effect size was significantly different from zero. If the 95% CI does not cross or touch the zero line, then you can say the effect was significant.

::: {.callout-note appearance="minimal"}
## Optional Additional Resources on Effect Sizes

-   How to use the `rstatix` package to calculate effect sizes

    -   **Cohen’s d for Student t-test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Cohen’s d for Welch test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Effect size** under **Wilcoxon rank sum test** on the [Wilcoxon Test in R](https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/) webpage

-   To learn more about inferential tests to compare two groups

    -   compare means with normal distribution: [t-Test](https://datatab.net/tutorial/t-test)

    -   rank-sum test to compare two independent samples that are not normally distributed: [Mann-Whitney U-Test](https://datatab.net/tutorial/mann-whitney-u-test)

    -   signed-rank test to compare paired groups that are not normally distributed: [Wilcoxon signed-rank test](https://datatab.net/tutorial/wilcoxon-test)

-   [Alternatives to P value: confidence interval and effect size](https://pmc.ncbi.nlm.nih.gov/articles/PMC5133225/)
:::

## Join Original Graph and Effect Size Results

Now let's make one image that joins your original graph with the effect size results graph together.

```{r}
#| label: data plus effect size graph

patch.plot <- best.plot + effect.plot +
  plot_layout(widths = c(3, 1))

patch.plot 

# save the graph
# recommended to use.png or .jpg file types
ggsave(filename = "results/datavizWithEffectSize.png",  
       plot = patch.plot,   #put the object name of your plot here
       height = 5, width = 8, units = "in", 
       dpi = 300)


```

The above code will save your new graph that includes the effect size results and call it "datavizWithEffectSize.png". This file will appear in the "results" subfolder of the dataviz folder under the Files tab in the lower right quadrant of this window.

If you are working on the Smith server, you will first need to export this file to your computer:

-   Check the box next to the "datavizWithEffectSize.png" file

-   Go to the "More" menu with the gear icon and click the drop down arrow

-   Select "Export" and then save the file to your computer

Once you have a PNG image file saved on your computer, replace the graph you made during the last lab in your Water Quality Results first draft with this new image. Make sure to update the figure legend of your new graph to add a description of the effect size that includes:

-   What type of standardized effect size test was it? (Cohen's d, Cohen's d for Welch test, or Wilcoxon effect size)

-   What does the green point represent?

-   What do the green vertical lines represent?

-   What does the horizontal dashed line represent?

## Render this File to PDF

You are now all done with this activity, but you want to make sure to save all your work and put a copy of it in your Individual Google Drive folder for this class (along side your lab notebook). Like in the last lab, you need a PDF file to put next to your lab notebook to keep a record of your data analysis.

-   Before you Render

    -   Make sure that each code chunk has an unique label name or no name

    -   Save this file (make sure this file is named "effectsize_YourName.qmd" (replace "YourName" with your full name)

-   Render this Quarto file to PDF (just select the Render menu item at the top of this window)

-   Make sure the PDF file is saved somewhere on your computer that you can get to (you may need to export this file to your computer if you are working on the Smith server)

-   Place a copy of the PDF file in your INDIVIDUAL Google Drive folder for this class
