---
title: "Intro to Data Visualization"
author: "Your Name Here"
date: "9/29/2020"
output: html_document
---

## Restart your R session 
Before getting started, it is a good practice to start with a clean slate by restarting your R session

Go to the top menu, select Session then select Restart R


## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.
It is recommended that you put all your code to load packages in one R chunk at the start

```{r load pacakges, eval=TRUE, include=FALSE}

library(tidyverse)
library(ggbeeswarm)
library(swirl)

```

## Import the Water Quality Data

```{r import data}

waterQ <- read_csv("waterQ.csv",
                   col_types = cols(date = col_date(format = "%m/%d/%Y"), 
                                    year = col_factor(),
                                    season = col_factor(levels = c("Summer", "Fall")),
                                    location = col_factor(levels =
                                              c("Downstream","Upstream")),
                                    microhabitat = col_factor(levels = 
                                              c("DSR", "DFR", "DM", "DSH", "DSP", 
                                                "USR", "UFR", "UM", "USH", "USU"))
                                    ))

```


Use the skills you learned in the "Getting Data into R" Swirl lesson to look at the structure of the dataframe that was just imported.  

I personally usually click on the name of the dataframe in the Global Environment as a shortcut to using the View() function, and click on the blue arrow next to the dataframe in the Global Environment to see all the variable names and types

Also click on the Files tab and click on the file called "waterQ_metadata.txt" to see a full description of the data including how to cite the data, what each variable represents, and how measurements were made using what tools.

## How does Paradise Pond affect water quality in the Mill River?

We can use this dataset to look at differences in the water quality upstream versus downstream of Paradise Pond to test for possible effects of the pond on the Mill River.

Paradise Pond is an impoundment in the river that slows the flow and makes it more like a pond and less like a river. The water falling over the dam creates a lot of turbulence.  How might the slow flow in the pond and/or the turbulence from the water falling over the dam affect the water quality downstream vs upstream?

**Question: Look at the 7 water quality variables and choose ONE that you want to explore today.  Type below which variable you chose:**




**Question: What do you predict to see in the data comparing the upstream and downstream locations for this variable? (in other words, How do you predict Paradise Pond will affect this variable (if at all)?).  Write your prediction and explain why you made that prediction below:**

For help understanding the variable better and how the pond might affect it, see the [Background Information on Water Quality Measurements](https://docs.google.com/document/d/1-rDncCF33sDK6o3ZdSo-zzSdCUxo77DIoDGmnp4i8EI/edit?usp=sharing)





## Missing Data

Does your variable of interest have missing data?  Run the code below to summarize the whole dataset

```{r Summary of Data}

summary(waterQ)

```

**Question: How many NA's (=missing data) does your variable of interest have?**



Let's remove the NA's for your variable.  Run the code below making sure to replace the ____ with the name of your variable of interest.

```{r remove NAs}

waterQ2 <- waterQ %>% 
  filter(!is.na(_____))

```

While this step is not strictly necessary, let's remove some of the columns we don't need to make the data easier to look at.  Run the code below making sure to replace the ____ with the name of your variable of interest.


```{r remove unneeded columns}

waterQ2 <- waterQ2 %>% 
  select(date, year, season, location, sampleID, sampling, microhabitat, sample_num,
         _______)

```


Look at the new dataset (waterQ2) in the Global Environment.  The new number of observations should be the original number minus the number of NA's.  The number of variables should be 9.


## Descriptive Statistics

Next we are going to calculate the descriptive statistics.  Before you do this for the Mill River water quality data, let's first get some practice by doing the "Descriptive Statistics" lesson from the "Intro to R for Biology" Swirl course.

Have one person from your group run the code below to start Swirl, then work in the console to run through the "Descriptive Statistics" lesson.  Go through the lesson together as a group.  When you finish the lesson, exit Swirl (if you at a command prompt, you can type bye() to exit Swirl.  You can select 0 or press escape if you are at a point where it asks you to select something).

See the "swirl-commands.txt" and "swirl-tips.txt" files for help with Swirl


```{r run swirl1, eval=FALSE, include=FALSE}

swirl()

```

Let's calculate the descriptive statistics for your variable of interest for the Mill River water quality data. Run the code below making sure to replace the ____ with the name of your variable of interest.


```{r descriptive stats}

waterQ2 %>% 
  group_by(location) %>% 
  summarize(mean = mean(_____), 
            median = median(_____), 
            SD = sd(_____), 
            IQR = IQR(_____), 
            n())

```
How do we report these values? We have to round appropriately and include the correct units.

**Question: What units is your variable of interest measured in? (Look at the "waterQ_metadata.txt" file in the data folder and then scroll down to your variable of interest to see what units it was measured in)**





How should we round these values? 

Let's start with the median.

If there is a odd number of values, then median is just the middle number of the dataset when arranged from smallest to largest value.  In this case, you should report the median in the same way the data were originally recorded.  Note, you can look at the "waterQ_metadata.txt" file in the data folder and then scroll down to your variable of interest to see that level of accuracy the variable was measured to. You can usually go one digit to the right beyond the level of accuracy (because that next digit is the estimated digit) at most.  Note that it is okay to be a little more conservative and round more, you just don't want to give the impression that you have more accuracy than you do by giving too many digits.

If there is an even number of values, then the median is the average of the two middle numbers.  Because calculating an average includes division, you can go one digit further to the right than how you reported the original values at most. 

Round the interquartile range (IQR) the same way you do the median.

**Question: Give the rounded values for median and IQR for each group below:**

Downstream median:
Downstream IQR:
Upstream median:
Upstream IQR:


Next let's round the standard deviation (SD).  Recall that you should round the SD to 1-2 significant digits at the most.

**Question: Give the rounded values for SD (use 2 significant digits) for each group below:**

Downstream SD:
Upstream SD:

To round the mean, you round it to the same DECIMAL place as the SD.  Rounding the SD tells us how variable the data are and how many decimals we can report to.

**Question: Give the rounded values for mean for each group below:**


Downstream mean:
Upstream mean:


**Question: Are the values for the mean and median for each group similar or different? (you have to use some judgment about what is "similar" vs "different" at this point)**



**Question: Can you tell at this point if the data for your variable of interest is normally distributed or not?**



## Data Visualization

Next we are going to visualize the data.  Before you do this for the Mill River water quality data, let's first get some practice by doing the "Graphing with ggplot" lesson from the "Intro to R for Biology" Swirl course.

Have one person from your group run the code below to start Swirl, then work in the console to run through the "Graphing with ggplot" lesson.  Go through the lesson together as a group.  When you finish the lesson, exit Swirl.


```{r run swirl2, eval=FALSE, include=FALSE}

swirl()

```

### Histogram

Let's first use a histogram to look at the distribution of your water quality variable of interest for each group.

The code to make histograms for each of the water quality variables is included in the code chunk below.  Notice that the only thing that differs is the x-variable, the binwidth, and the x-label.

Delete all the code except for your variable of interest, then run the code to make a histogram for your variable. 


```{r Histogram, eval=TRUE, include=TRUE}


#Histogram for Flow 
ggplot(data = waterQ2, aes(x = flow_rate_m.s, fill=location))+ 
  geom_histogram(binwidth = 0.05, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab("Flow Rate (m/s)") +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20) 

#Histogram for DO
ggplot(data = waterQ2, aes(x = DO_mg.L, fill=location))+ 
  geom_histogram(binwidth = 0.1, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab("Dissolved Oxygen (mg/L)") +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)

#Histogram for Light
ggplot(data = waterQ2, aes(x = light_umol.m2.s, fill=location))+ 
  geom_histogram(binwidth = 50, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab(~paste("PAR"," ", "(" ,mu, "mol m"^-2,"s"^-1, ")")) +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)

#Histogram for pH
ggplot(data = waterQ2, aes(x = pH, fill=location))+ 
  geom_histogram(binwidth = 0.05, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab("pH") +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)

#Histogram for Water Temperature
ggplot(data = waterQ2, aes(x = water_temp_C, fill=location))+ 
  geom_histogram(binwidth = 0.5, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab("Water Temperature ("~degree~"C)") +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)

#Histogram for Turbidity
ggplot(data = waterQ2, aes(x = turbidity_NTU, fill=location))+ 
  geom_histogram(binwidth = 1, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab("Turbidity (NTU)") +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)

#Histogram for Conductivity    
ggplot(data = waterQ2, aes(x = conductivity_uS.cm, fill=location))+ 
  geom_histogram(binwidth = 5, color = "white", show.legend = FALSE) +
  facet_grid(location ~ .) +
  xlab(~paste("Conductivity"," ", "(" ,mu, "S cm"^-1, ")")) +
  ylab("Count") +
  coord_cartesian(expand=FALSE) +
  theme_classic(base_size=20)


```




**Question: Does the distribution for each location look like a normal distribution?**



### Choosing the Best Graph

The last thing we will do is choose the best way to graph your data to present in a paper or presentation.  Go to the code from the [Graphing Grouped Continuous Data Swirl lesson](https://github.com/marneypratt/introR4bio/blob/master/Graphing_Grouped_Continuous_Data-code.md) and copy the code of the graph type you think would be best in the code chunk below (feel free to try several before you decide which one is best).  

Tweak the template code from the Swirl lesson so that it works with your water quality variable of interest.  Feel free to play around to see what happens when you remove or change things in the code.


```{r best graph}



```








## Exporting your best graph

The simplest thing to do is probably just right-click (Windows) or control-click (Mac) on the graph and select "Copy Image" and then just paste the image where you want it.  However, if you want a higher quality image where you can specify the size, image quality, and file type you can use the ggsave function.

When saving with the ggsave function, you need to name the resulting file with surrounding “ “, tell R which ggplot object we want (assign the plot to a name such as g1), and indicate the size via height, width, and units. Don't forget to save the graph with a dpi call to make it nice and crisp! 

```{r Save your graph, eval = F}

# save the graph!
ggsave(filename="graph-name.png", 
       plot=g1, 
       height = 7, width = 8, units = "in", 
       dpi = 500)

```
