---
title: "Intro to Data Visualization"
author: "Your Name Here"
date: "9/29/2020"
output: html_document
---

## Restart your R session 
Before getting started, it is a good practice to start with a clean slate by restarting your R session

Go to the top menu, select Session then select Restart R


## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.
It is recommended that you put all your code to load pacakges in one R chunk at the start

```{r load pacakges}

library(tidyverse)
library(ggbeeswarm)
library(swirl)

```

## Import the Water Quality Data

```{r import data}

waterQ <- read_csv("data/waterQ.csv",
                   col_types = cols(date = col_date(format = "%m/%d/%Y"), 
                                    year = col_factor(),
                                    season = col_factor(levels = c("Summer", "Fall")),
                                    location = col_factor(levels = c("Downstream","Upstream")),
                                    microhabitat = col_factor(levels = 
                                              c("DSR", "DFR", "DM", "DSH", "DSP", 
                                                "USR", "UFR", "UM", "USH", "USU"))
                                    ))

```


Use the skills you learned in the "Getting Data into R" Swirl lesson to look at the structure of the dataframe that was just imported.  

I personally usually click on the name of the dataframe in the Global Environment as a shortcut to using the View() function, and click on the blue arrow next to the dataframe in the Global Environment to see all the variable names and types
